Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-11-27 manuscript/manuscript.tex
Change introduction and methodology in manuscript

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting adaptive discretization algorithm that reduces
the errors due to the distance between the tesseroid and the computation
point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear and
exponential density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential, its
gradient, and Marussi tensor are 1, 2 and 8, respectively.
A delta ratio of 0.2 is needed for the computation of the gravitational potential and
its gradient components, while a value of 0.01 must be used for the Marussi tensor
components.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, its gradient and Marussi tensor components
by comparing the numerical model with the fields generated by a spherical shell.

@@

We present a new algorithm for computing the gravitational fields generated by any
tesseroid with any arbitrary continuous density function on any external point.
It is based on the GLQ approximation and the adaptive discretization of
\citet{Uieda2016}, which we extend to include a new density-based discretization step.
In order to ensure the accuracy of the numerical approximation we have
determined optimal values for the controlling parameters by
comparing the numerical approximation with analytical solutions for
spherical shells.
Finally, we applied the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with an exponentially increasing density with depth.

@@

\subsection{Adaptive Discretization}

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm, which we will use
here.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

The following algorithm computes the gravitational fields at a given point at $(r, \phi,
\lambda)$ due to a tesseroid with its geometric centre at $(r_t, \phi_t, \lambda_t)$ and
dimensions $L_r$, $L_\phi$, and $L_\lambda$ given by

@@

\begin{equation}
    L_\phi = r_2 \arccos(\sin\phi_2\sin\phi_1 + \cos\phi_2\cos\phi_1),
\end{equation}

and

\begin{equation}
    L_r = r_2 - r_1.
    \label{eq:sizer}
\end{equation}

\textit{Step 1}: Check that the tesseroid satisfies the following inequality for each
dimension $L_i$ of the tesseroid:

@@

Notice that the distance-size ratio $D$ determines how many
times the tesseroids will be divided.
Therefore, it effectively regulates both the accuracy of the algorithm and its
computation time.
An optimal value for $D$ cannot be directly calculated from the desired accuracy level.
Instead, it is empirically determined by comparing the numerical results with the
analytical solution for a spherical shell.
\citet{Uieda2016} used a shell with homogeneous density to determine optimal values of
$D$.
Here, we will repeat the numerical experiment using analytical expressions for shells
with density varying according to exponential and linear functions of $r$.

@@

The numerical integration of an arbitrary continuous density function introduces
a new type of problem:
the integration error from using only a few nodes to discretize the density function.
The adaptive discretization may help
to reduce this kind of error.
However, it does not take into account the density function
and hence it is not well suited to fully perform this task.

@@

\noindent
in which $L_r$ is the radial dimension of the tesseroid being considered for division,
$L_r^\text{orig}$ is the radial dimension of the original tesseroid, and
$\delta$ is a positive constant henceforth called the \textit{delta ratio}.

@@

>>>>

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear and
exponential density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential, its
gradient, and Marussi tensor are 1, 2 and 8, respectively.
A delta ratio of 0.2 is needed for the computation of the gravitational potential and
its gradient components, while a value of 0.01 must be used for the Marussi tensor
components.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000 m, using an exponential density function.
\end{summary}

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the three dimensional adaptive discretization algorithm of
\citet{Li2011} to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, its gradient and Marussi tensor components
by comparing the numerical model with the fields generated by a spherical shell.

@@

We present a new algorithm for computing the gravitational fields generated by any
tesseroid with any arbitrary continuous density function on any external point.
It is based on the GLQ approximation and a two dimensional version of the adaptive
discretization of \citet{Uieda2016}, which we extend to include a new density-based
discretization algorithm.
In order to ensure the accuracy of the numerical approximation we have
determined optimal values for the controlling parameters by
comparing the numerical approximation with analytical solutions for
spherical shells.
Finally, we applied the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with an exponentially increasing density with depth.

@@

\subsection{Two Dimensional Adaptive Discretization}

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm that speeds the
computation of the distances.
Both algorithms (the original one from \citet{Li2011} and the modified
version of \citet{Uieda2016}) performs tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2018} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the splitting makes the computation more efficient, while
keeping an acceptable accuracy on the gravitational fields \citep{Lin2018}.

Here we will use a two dimensional version of the adaptive discretization algorithm
developed by \citet{Uieda2016}.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

Lets consider a computation point located at $(r, \phi, \lambda)$ and an arbitrary
tesseroid with its geometric centre at $(r_t, \phi_t, \lambda_t)$.
We can define its longitudinal ($L_\lambda$) and latitudinal ($L_\phi$) dimensions as
follows:

@@

\begin{equation}
    L_\phi = r_2 \arccos(\sin\phi_2\sin\phi_1 + \cos\phi_2\cos\phi_1).
\end{equation}

\textit{Step 1}: Check that the tesseroid satisfies the following inequality for each
dimension $L_i (i \in \{\lambda, \phi\})$ of the tesseroid:

@@

Notice that the distance-size ratio $D$ determines how many
times the tesseroids will be divided.
Therefore, it effectively regulates both the accuracy of the algorithm and its
computation time.
An optimal value for $D$ cannot be directly calculated from the desired accuracy level.
Instead, it is empirically determined by comparing the numerical results with the
analytical solution for a spherical shell.
\citet{Uieda2016} used a shell with homogeneous density to determine optimal values of
$D$.
Here, we will repeat the numerical experiment using analytical expressions for shells
with density varying according to exponential and linear functions of $r$.
This numerical experiment will also show that the two dimensional adaptive
discretization algorithm is as accurate as the original three dimensional version,
though it reduces the number of tesseroid subdivisions and thus the computation time.

@@

The numerical integration of an arbitrary continuous density function introduces
a new type of problem:
the integration error from using only a few nodes to discretize the density function.
The three dimensional adaptive discretization may help
to reduce this kind of error by adding more point masses in the radial direction.
However, it does not take into account the density function
and hence it is not well suited to fully perform this task.
Moreover, because the two dimensional adaptive discretization algorithm does not
subdivide the tesseroid on the radial dimension, it does not put any effort on lowering
this new kind of error.

@@

\noindent
in which $L_r$ is the radial dimension of the tesseroid being considered for division,

\begin{equation}
    L_r = r_2 - r_1,
\end{equation}

\noindent $L_r^\text{orig}$ is the radial dimension of the original tesseroid, and
$\delta$ is a positive constant henceforth called the \textit{delta ratio}.

@@

\subsection{Algorithm summary}

In summary, given any tesseroid with variable density in depth and an arbitrary
external computation point, we propose the following steps to numerically
compute the gravitational fields:

\textit{Step 1:}
Apply the density-based discretization algorithm.
It subdivides the tesseroid only in the radial dimension, thus it will return a set of
tesseroids with the same longitudinal and latitudinal dimensions as the original one but
with different radial boundaries.

\textit{Step 2:}
Apply the two dimensional adaptive discretization algorithm on each tesseroid obtained
in the previous step.
If needed, it will subdivide each tesseroid only in the latitudinal and longitudinal
direction, generating a set of smaller tesseroids that resemble the original one.

\textit{Step 3:}
Apply a second-order GLQ to numerically compute the gravitational fields
(Eq.\ref{eq:glq-var-dens}) generated by each tesseroid of the set obtained in the
previous step.
Then sum these results in order to obtain the gravitational field of the original
tesseroid.
